Verifying Spatial Properties of Array Computations

### 2017-04-17 oopsla17/algorithms.tex

In the model, we can regard both array indices and specifications as groups of
points in $N$-dimensional space. Notion of consistency is then intuitively
whether these two (potentially infinite) sets of points in space match or not.
This leads to a simple set based consistency function defined as
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \textit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \textit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not.Then it delegates to $\mathit{consistent'}$
function which checks if the points observed in the array expression matches
the space defined by the specification. Lower bounds, marked by \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost} modifier,
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices correspond exactly those defined by the
region in the specification, hence requiring set equality.

Reader will realise that since index schemes allows infinite intervals in the
product, set equality cannot be done by exhaustively checking the points of one
are inside the other. Instead, we compile the region into interval constraints
and subscript schemes into membership constraints and pass it to
\textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The query
belongs to to quantifier free linear arithmetic, which is a decidable theory.
Performance of consistency checking using SMT solver is irrelevant as the number
of interval constraints and the number of indices that can reasonably appear in
an assignment is too small restricting the size of the formula being checked.

\mnote{It would be great if we had data on number of indexing expressions that
occur (counting the ones that flow in) and also the length of specifications
(which I think we do generate). This would justify my efficiency claims}.

>>>>

In the model, we can regard both array indices and specifications as groups of
points in $N$-dimensional space. Notion of consistency is then intuitively
whether these two (potentially infinite) sets of points in space match or not.
This leads to a simple set based consistency function defined as
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
function to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost} modifier,
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices correspond exactly those defined by the
region in the specification, hence requiring set equality.

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite and set comparisons cannot be done by exhaustively
comparing the points of one with the other. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
it to \textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The
query belongs to to quantifier free linear arithmetic, which is a decidable
theory.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}. Performance of consistency checking
using satisfiability is irrelevant as the the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions conntected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
99\% of all array variables that we can assign a specification is X, and 99\% of
number of array terms that flows into such an assignment is Y. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

\mnote{DOMINIC fills in X, Y, and Z}
